---
title: "Report EDA"
author: "Christian Pascual"
output: pdf_document
---

```{r, message = F, warning = F }
library(tidyverse)
set.seed(1)

# Load the data and do some processing
cancer = read_csv("cancer_registry.csv") %>% 
  # Split the geography variable
  separate(Geography, into = c("county", "state"), sep = ", ") %>% 
  # Split up binnedInc into a lower and upper decile
  mutate(
    binnedInc = str_remove_all(binnedInc, "[(\\]]"),
    # also try to group states by region
    region = case_when(
      state %in% c("California", "Oregon", "Washington", "Nevada", "Idaho", 
                   "Montana", "Wyoming", "Colorado", "Utah", "Alaska", "Hawaii") ~ "West",
      state %in% c("Arizona", "New Mexico", "Texas", "Oklahoma") ~ "Southwest",
      state %in% c("North Dakota", "South Dakota", "Nebraska", "Kansas", 
                   "Minnesota", "Iowa", "Missouri", "Wisconsin", "Illinois", 
                   "Indiana", "Ohio", "Michigan") ~ "Midwest",
      state %in% c("Arkansas", "Louisiana", "Mississippi", "Alabama", "Georgia",
                   "Florida", "South Carolina", "North Carolina", "Tennessee",
                   "Kentucky", "Virginia", "West Virginia", "District of Columbia",
                   "Delaware") ~ "Southeast",
      state %in% c("Maryland", "Pennsylvania", "New Jersey", "New York", "Rhode Island",
                   "Connecticut", "Massachusetts", "New Hampshire", "Vermont", "Maine") ~ "Northeast",
      TRUE ~ "Southwest" # Weird formatting means a single NM is NA in state
    )
  ) %>% 
  separate(binnedInc, into = c("inc_dec_low", "inc_dec_high"), sep = ",") %>% 
  janitor::clean_names() %>%  # Convert all column names to lowercase'
  mutate(
    high_college = pct_bach_deg25_over > median(pct_bach_deg25_over), # median(pct_bach_deg18_24),
    high_hs = pct_hs18_24 > median(pct_hs18_24)
  )
```

# Missing Data

```{r}
# Get the number of rows with missing data for each column
cancer %>% 
  select(everything()) %>%
  summarise_all(funs(sum(is.na(.)))) %>% 
  select(which(colMeans(.) != 0))
```

Three columns have missing data:

- `pct_some_col18_24`
- `pct_employed16_over`
- `pct_private_coverage_alone`

# Distribution of outcome

```{r}
cancer %>% 
  ggplot(aes(x = target_death_rate)) +
  geom_histogram(color = "red") +
  ggtitle("Empirical distribution of cancer death rate in the data")
```

Outcome is reasonably normally distributed, which is nice. 

# Variable Trends

## Incidence and cancer death rate

```{r}
cancer %>% 
  ggplot(aes(x = incidence_rate, y = target_death_rate)) +
  geom_point() +
  ggtitle("Incidence of cancer vs death rate")
```

Suggests a positive relationship between higher incidence and higher death rate. This is expected, so this should be controlled for in the model.

## Age and death rate

```{r}
cancer %>% 
  filter(median_age < 100) %>% # some weird values in the data
  ggplot(aes(x = median_age, y = target_death_rate)) +
  geom_point() +
  ggtitle("Age vs death rate")
```

I thought there might be a relationship between age and death rate. Doesn't seem to show here. I know that the older you get, the more likely you are to get cancer, so this should still be controlled for in the analysis.

## Education & Death Rate

```{r}
cancer %>% 
  pivot_longer(
    cols = c("pct_no_hs18_24", "pct_hs18_24", "pct_bach_deg18_24"),
    values_to = "pct",
    names_to = "educ_level"
  ) %>% 
  mutate(
    educ_level = case_when(
      educ_level == "pct_no_hs18_24" ~ "No HS", 
      educ_level == "pct_hs18_24" ~ "HS", 
      educ_level == "pct_bach_deg18_24" ~ "College"
    )
  ) %>% 
  ggplot(aes(x = pct, y = target_death_rate, color = educ_level)) + 
  facet_grid(educ_level ~ .) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm") +
  labs(
    title = "Cancer mortality by percent of county with particular education level",
    x = "Percent of population with particular education level",
    y = "Cancer Mortality (deaths / 100K)")
```

Hard to parse much here. It kind of looks like more education leads to slightly lower death rates (based on blue plot)? But this effect is not very pronounced.

## Insurance Coverage & Death Rate

```{r}
cancer %>% 
  pivot_longer(
    cols = c("pct_private_coverage", 
             "pct_private_coverage_alone", 
             "pct_public_coverage",
             "pct_public_coverage_alone"),
    values_to = "pct",
    names_to = "ins_level"
  ) %>% 
  ggplot(aes(x = pct, y = target_death_rate, color = ins_level)) + 
  facet_grid(ins_level ~ .) +
  theme(legend.position = "bottom") +
  geom_point() +
  ggtitle("Death rate by percent of county by percentage of people with insurance types")
```

Interesting shift. The higher private coverage is, the lower the death rate. It's converse for public coverage. Perhaps this is a proxy for other economic factors? Maybe private insurance is better than public coverage? My thought was that more coverage in general would help improve death rate.

## Race & Death Rate

```{r}
cancer %>% 
  pivot_longer(
    cols = c("pct_white", 
             "pct_black", 
             "pct_asian",
             "pct_other_race"),
    values_to = "pct",
    names_to = "race"
  ) %>% 
  mutate(
    race = case_when(
      race == "pct_white" ~ "White",
      race == "pct_black" ~ "Black",
      race == "pct_asian" ~ "Asian",
      race == "pct_other_race" ~ "Other"
    )
  ) %>% 
  ggplot(aes(x = pct, y = target_death_rate, color = race)) + 
  geom_point(size = 0.5, alpha = 0.3) +
  geom_smooth(method = "lm") +
  facet_grid(race ~ .) +
  theme(legend.position = "bottom") +
  labs(
    title = "Death rate by percent of county by percentage of race in a county",
    x = "Percentage of race in county population",
    y = "Cancer Mortality (deaths / 100K)")
```

Main takeaway here is that a higher percentage of African-Americans corresponds to higher death rates. This lines up with evidence of racial disparities in the healthcare system with regard to access and cost. 

## Region by Death Rate

```{r}
cancer %>% 
  ggplot(aes(x = region, y = target_death_rate, color = region)) + 
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "Death rate by US region",
    x = "Region",
    y = "Cancer Mortality (deaths / 100K)") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  )
```

Interesting... Southeast seems to have a higher death rate than the other regions. 

# Interaction Between Education and Race?

```{r}
cancer %>% 
  mutate(
    high_college = pct_bach_deg25_over > 34, # median(pct_bach_deg18_24),
    high_hs = pct_hs18_24 > median(pct_hs18_24)
  ) %>% 
  pivot_longer(
    cols = c("pct_white", 
             "pct_black", 
             "pct_asian",
             "pct_other_race"),
    values_to = "pct",
    names_to = "race"
  ) %>% 
  mutate(
    race = case_when(
      race == "pct_white" ~ "White",
      race == "pct_black" ~ "Black",
      race == "pct_asian" ~ "Asian",
      race == "pct_other_race" ~ "Other"
    )
  ) %>% 
  ggplot(aes(x = pct, y = target_death_rate, color = high_college)) + 
  geom_smooth(method = "lm", size = 0.5) +
  facet_grid(race ~ .) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5)
  ) + 
  labs(
    title = "Death rate by percent of by race in a county and high education",
    x = "Percentage of race in county population",
    y = "Cancer Mortality (deaths / 100K)") +
  scale_color_discrete(labels = c("< 50%", ">= 50%"))
```

```{r}
cancer %>% 
  ggplot(aes(x = pct_bach_deg25_over)) +
  geom_density()
```

```{r}
cancer %>% 
  ggplot(aes(x = pct_hs25_over)) +
  geom_density()
```

# Inference Model

```{r}
colnames(cancer)
```

```{r}
# Trying out fuill interaction
model = lm(target_death_rate ~ pct_no_hs18_24 + pct_no_hs18_24 + pct_some_col18_24 +
                               pct_bach_deg18_24 + pct_white + pct_black + pct_asian + pct_other_race +
                               pct_no_hs18_24*pct_white + pct_no_hs18_24*pct_black + pct_no_hs18_24*pct_asian + 
                               pct_no_hs18_24*pct_other_race + pct_hs18_24*pct_white + pct_hs18_24*pct_black + pct_hs18_24*pct_asian + 
                               pct_hs18_24*pct_other_race + pct_some_col18_24*pct_white + pct_some_col18_24*pct_black + pct_some_col18_24*pct_asian + 
                               pct_some_col18_24*pct_other_race + pct_bach_deg18_24*pct_white + pct_bach_deg18_24*pct_black + 
                               pct_bach_deg18_24*pct_asian + pct_bach_deg18_24*pct_other_race +
                               incidence_rate + med_income + pop_est2015 + poverty_percent + median_age,
                             data = cancer)

# Doesn't seem very good, will try indicators for high education and low education
summary(model)             
```

```{r}
# Trying out different set of education vars
model2 = lm(target_death_rate ~ pct_hs25_over + pct_bach_deg25_over + 
             pct_white + pct_black + pct_asian + pct_other_race +
             # Interactions
             pct_white*pct_hs25_over + pct_black*pct_hs25_over + pct_asian*pct_hs25_over + pct_other_race*pct_hs25_over +
             pct_white*pct_bach_deg25_over + pct_black*pct_bach_deg25_over + pct_asian*pct_bach_deg25_over + pct_other_race*pct_bach_deg25_over +
             # Confounders
             incidence_rate + med_income + pop_est2015 + poverty_percent + median_age,
           data = cancer)

# Looks good! Some interesting results here.
summary(model2)      
```


```{r}
# Function for recreating the interaction model for bootstrapping
create_int_model = function(data) {
  
  # lm(target_death_rate ~ pct_no_hs18_24 + pct_no_hs18_24 + pct_some_col18_24 +
  #            pct_bach_deg18_24 + pct_white + pct_black + pct_asian + pct_other_race +
  #            pct_no_hs18_24*pct_white + pct_no_hs18_24*pct_black + pct_no_hs18_24*pct_asian + 
  #            pct_no_hs18_24*pct_other_race + pct_hs18_24*pct_white + pct_hs18_24*pct_black + pct_hs18_24*pct_asian + 
  #            pct_hs18_24*pct_other_race + pct_some_col18_24*pct_white + pct_some_col18_24*pct_black + pct_some_col18_24*pct_asian + 
  #            pct_some_col18_24*pct_other_race + pct_bach_deg18_24*pct_white + pct_bach_deg18_24*pct_black + 
  #            pct_bach_deg18_24*pct_asian + pct_bach_deg18_24*pct_other_race +
  #            incidence_rate + med_income + pop_est2015 + poverty_percent + median_age,
  #          data = data)
  
  
  lm(target_death_rate ~ pct_hs25_over + pct_bach_deg25_over + 
             pct_white + pct_black + pct_asian + pct_other_race +
             # Interactions
             pct_white*pct_hs25_over + pct_black*pct_hs25_over + pct_asian*pct_hs25_over + pct_other_race*pct_hs25_over +
             pct_white*pct_bach_deg25_over + pct_black*pct_bach_deg25_over + pct_asian*pct_bach_deg25_over + pct_other_race*pct_bach_deg25_over +
             # Confounders
             incidence_rate + med_income + pop_est2015 + poverty_percent + median_age,
           data = data)
  
}
```

```{r}
# How many bootstrap datasets do I want
bs_n = 5000

# Terms to keep
# terms = model$coefficients %>% names %>% .[15:30]
terms = model2$coefficients %>% names %>% .[13:20]

# Create the bootstrap datasets and models
bs = tibble( idx = 1:bs_n ) %>% 
  mutate(
    bs_data = map(idx, function(i) {
      sample_n(cancer, size = nrow(cancer), replace = TRUE)
    }),
    bs_model = map(bs_data, function(bsd) {
      create_int_model(bsd)
    }),
    bs_results = map(bs_model, broom::tidy)
  ) %>% 
  select(idx, bs_results) %>% 
  unnest(bs_results) %>% 
  group_by(term) %>% 
  summarize(
    n = n(),
    bs_mean = mean(estimate),
    bs_var = var(estimate),
    left_bound = quantile(estimate, 0.025),
    right_bound = quantile(estimate, 0.975),
  ) %>% 
  filter(term %in% terms) %>% 
  # Convert terms to factors for easier reordering
  mutate(
    # Model 1 formatting
    # term = factor(term, 
    #               levels = c(
    #                 "pct_no_hs18_24:pct_asian", "pct_asian:pct_hs18_24",
    #                 "pct_some_col18_24:pct_asian", "pct_bach_deg18_24:pct_asian",
    #                 "pct_no_hs18_24:pct_black", "pct_black:pct_hs18_24",
    #                 "pct_some_col18_24:pct_black","pct_bach_deg18_24:pct_black",
    #                 "pct_no_hs18_24:pct_other_race", "pct_other_race:pct_hs18_24",
    #                 "pct_some_col18_24:pct_other_race", "pct_bach_deg18_24:pct_other_race",
    #                 "pct_no_hs18_24:pct_white", "pct_white:pct_hs18_24",   
    #                 "pct_some_col18_24:pct_white", "pct_bach_deg18_24:pct_white"),
    #               labels = c(
    #                 "No HS x Asian", "Some HS x Asian",
    #                 "Some College x Asian", "College x Asian",
    #                 "No HS x Black", "Some HS x Black",
    #                 "Some College x Black", "College x Black",
    #                 "No HS x Other", "Some HS x Other",
    #                 "Some College x Other", "College x Other",
    #                 "No HS x White", "Some HS x White",
    #                 "Some College x White", "College x White"
    #               ))
    term = factor(term, 
                  levels = c(
                    "pct_hs25_over:pct_asian", "pct_bach_deg25_over:pct_asian",
                    "pct_hs25_over:pct_black", "pct_bach_deg25_over:pct_black",
                    "pct_hs25_over:pct_other_race", "pct_bach_deg25_over:pct_other_race",
                    "pct_hs25_over:pct_white", "pct_bach_deg25_over:pct_white"),
                  labels = c(
                    "HS x Asian", "College x Asian",
                    "HS x Black", "College x Black",
                    "HS x Other", "College x Other",
                    "HS x White", "College x White"
                  ))
  )

# Visualize the bootstrap confidence intervals
bs %>% 
  ggplot(aes(x = term, y = bs_mean)) +
  geom_pointrange(aes(ymin = left_bound, ymax = right_bound,
                      color = if_else(left_bound > 0 | right_bound < 0, "y", "n"))
                  ) +
  geom_hline(yintercept = 0, color = "red", alpha = 0.5) +
  coord_flip() +
  labs(
    title = "Bootstrap confidence intervals for 5000 resamples",
    x = "Interaction Term",
    y = "Bootstrap Interaction Coefficient Estimate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"
  ) +
  scale_color_manual(values = c("#DC143C", "#2E8B57"))
```





